Technical Paper Proposal: \\ Placement Algorithms for Heterogenous FPGAs

### 2024-09-11 docs/proposal/proposal.tex
more prop
remove comments
more waffling

\section{Proposal}

% In the software world, an engineer writes a program in a high-level langauge like C, Java, or Python. 
% Then the engineer compiles the program using some compilation software like gcc or clang. 
% First, the compiler takes the engineer's high-level program, and through multiple stages of lexing, parsing, 
% and optimizations, generates a low-level program in an assembly language like x86 or MIPS. 
% Then, the compiler assembles the assembly code into machine code. 
% The produced machine code is often referred to as a binary and is ready to execute on the machine. 
% Different compilers may have different strategies for this process. 
% For example, LLVM and clang skip the assembly stage and go directly from high-level code to machine code, while gcc sticks to the traditional two-step process. 
% The Java and Python are more complex in that they compile high-level code into an intermediate language as bytecode. 
% In Java, the compiler assembles the bytecode into machine code at runtime in a process called Just-in-Time compilation which allows it to be platform independent. 
% Python offers the choice to assemble the machine code before or at runtime. 
% Simplify this downto 2 sentences.

In modern software development, the process of compiling high-level languages like C++ or Java into executable machine code is highly efficient and streamlined. 
The software toolchains only consist of compilers and scripts and are mostly open source and easy to learn. 
The compilation is where most of the magic happens.
Through multiple stages of lexing, parsing, and optimizations, the compiler takes the developer's program and produces working machine code.
The developer can influence a few optimization flags if needed, but for the most part, the developer simply runs the default compiler command and is presented with the final product.
All of this happens within milliseconds to seconds, even for large projects.

@@

Dramatic transition.

In the FPGA world, the design entry to product pipeline is despairingly complex by comparison. 
First, the design entry. The engineer describes a digital system using a high-level HDL like Verilog or VHDL. 
Then the engineer submits the design entry to an EDA like Vivado or Quartus to perform three stages of automated design. 
These are: synthesis, placement, and routing. 
All three are NP-hard problems.
For even modest designs like binary counters or 8-bit CPUs that utilize a small percentage of the available resources on the FPGA, the EDA will spend a minimum of a couple minutes building the design.
The EDA build time increases exponentially with the scale of the project. 
Projects that utilize 80\% or more of the FPGA's resources may build for hours, and for the high-end devices like Xilinx's Kintex which boast millions of logical elements, can take days.
Sometimes the EDA cannot fit the design onto the device.
Sometimes you step out of the doll-house and the world seems like a heironymous bosch painting.

\section{Abbreviations}
\begin{itemize}[label={--}, left=0.25cm] % Adjust label and left indent as desired
    \item \textbf{FPGA}: Field Programmable Gate Array
    \item \textbf{VLSI}: Very Large Scale Integration
    \item \textbf{EDA}: Electronic Design Automation
    \item \textbf{VHSIC}: Very High Speed Integrated Circuits
    \item \textbf{HDL}: Hardware Description Language
    \item \textbf{VHDL}: VHSIC HDL
    \item \textbf{HLS}: High Level Synthesis: Generating synthesizable HDL from high-level software languages. A company might want to have a software engineer write C or C++ code and have a program translate it into synthesizable Verilog. This can can boost productivity and save the company the need to hire a hardware engineer.
    \item \textbf{IP}: Intellectual Property: In FPGA context, this means pre-built modules or subsystems like a hardened microprocessor or Ethernet controller. These are usually proprietary.
    \item \textbf{SoC}: System on Chip: An FPGA device (chip) that features hardened IP in addition to the programmable logic fabric.
    \item \textbf{PL-PS}: Programmable Logic - Processing System: A design that utilizes the on-chip hard microprocessor in conjunction with the programmable logic fabric.
    \item \textbf{EDIF}: Electronic Design Interchange Format
    \item \textbf{HPWL}: Half Perimeter Wire Length
\end{itemize}

>>>>

\section{Proposal}
In modern software development, the process of compiling high-level languages like C++ or Java into executable machine code is highly efficient and streamlined. 
The software toolchains only consist of compilers and scripts and are mostly open source and easy to learn. 
The compilation is where most of the magic happens.
Through multiple stages of lexing, parsing, and optimizations, the compiler takes the developer's program and produces working machine code.
The developer can influence a few optimization flags if needed, but for the most part, the developer simply runs the default compiler command and is presented with the final product.
All of this happens within milliseconds to seconds, even for large projects.

@@

In the FPGA world, the design entry to product pipeline is despairingly complex by comparison. 
First, the design entry. The engineer describes a digital system using a high-level HDL like Verilog or VHDL. 
Then the engineer submits the design entry to an EDA like Vivado or Quartus to perform three stages of automated design. 
These are: synthesis, placement, and routing. 
All three are NP-hard problems.
For even modest designs like binary counters or 8-bit CPUs that utilize a small percentage of the available resources on the FPGA, the EDA will spend a minimum of a couple minutes synthesizing and implementing the design.
The EDA run time increases exponentially with the scale of the project. 
Projects that utilize 80\% or more of the FPGA's resources may run for hours, and for the high-end devices like Xilinx's Kintex FPGAs which have millions of logical elements, can run for days.
At high utilization, sometimes the design EDA cannot fit the design onto the device due to routing congestion even when it has enough logical elements to implement the design, essentially wasting hours or days of runtime.

One of the most common complaints from those in the FPGA design industry, especially from those in software development, is that the FPGA toolchains are overall slow and heavy. 
Unlike in software development where an error in the program can be patched and recompiled in a matter of minutes, EDA runtimes lasting several hours means that an engineer can only go through a couple design cycles per day. 
A Verilog developer must be very precise with their coding and practice thorough verification of their design before submitting it to the EDA.
This overall heightens the barrier of entry into FPGA development and contributes to a shortage of qualified FPGA engineers.

This paper aims to attack one of the sections of this wall: the placement stage. Like synthesis and routing, it is an NP-hard problem. The search space for a placed design is massive.

\section{Abbreviations}
\begin{itemize}[label={--}, left=0.25cm]
    \item \textbf{FPGA}: Field Programmable Gate Array
    \item \textbf{VLSI}: Very Large Scale Integration
    \item \textbf{EDA}: Electronic Design Automation
    \item \textbf{VHSIC}: Very High Speed Integrated Circuits
    \item \textbf{HDL}: Hardware Description Language
    \item \textbf{VHDL}: VHSIC HDL
    \item \textbf{HLS}: High Level Synthesis: Generating synthesizable HDL from high-level software languages. A company might want to have a software engineer write C or C++ code and have a program translate it into synthesizable Verilog. This can can boost productivity and save the company the need to hire a hardware engineer.
    \item \textbf{IP}: Intellectual Property: In FPGA context, this means pre-built modules or subsystems like a hardened microprocessor or Ethernet controller. These are usually proprietary.
    \item \textbf{SoC}: System on Chip: An FPGA device (chip) that features hardened IP in addition to the programmable logic fabric.
    \item \textbf{PL-PS}: Programmable Logic - Processing System: A design that utilizes the on-chip hard microprocessor in conjunction with the programmable logic fabric.
    \item \textbf{EDIF}: Electronic Design Interchange Format
    \item \textbf{HPWL}: Half Perimeter Wire Length
\end{itemize}
